A tutorial on the model-dependent analysis of neutron and X-ray reflectometry

### 2020-08-05 refl_maths/paper.tex

\begin{abstract}
Neutron reflectometry analysis is an inherently ill-posed, which is to say that there are many possible solutions which agree equally well with the measured data. 
This leads to the use of model-dependent analysis to interpret the experimental data, where information that is ``known'' about the system is integrated into the analysis.
This tutorial aims to introduce the mathematics that underlies the use of model-dependent analysis in neutron reflectometry. 
I hope that those that are well experienced in reflectometry find it as useful as those who are just about to experience their first reflectometry experiment. 
\end{abstract}

@@

The Born approximation \cite{born_quantenmechanik_1926}, which assumes a single scattering event for each incident radiation particle, allows the reflected intensity of be calculated as follows \cite{sivia_elementary_2011},
%
\begin{equation}
    R(q) \approx \frac{16\pi^2}{q^4} \bigg| \int^{+\infty}_{-\infty}{\rho'(z)\exp{(-\mathrm{i} zq) \text{d}z} \bigg|^2},
    \label{equ:kine}
\end{equation}
%
where, $\rho'(z)$ is the first derivative of the scattering length density profile.
The scattering length density profile is our model, a functional description of how the scattering length density varies with depth normal to the interface, $z$.

@@

\section{Recursive methods}
The breakdown of the kinematic approach leads to the application of the Abel\`{e}s \cite{abeles_sur_1948} or Parratt \cite{parratt_surface_1954}, recursive model for the reflection of light at a given number of stratified interfaces, also referred to as the dynamical approach. 
This is an alternative methods for calculating the reflected intensity from a model, however, now the model is a series of layers of different scattering length density. 
The probing radiation can be either reflected or refracted at the interface between each of the layers. 
%
\begin{figure}[t]
    \includegraphics[width=\linewidth]{dynamic}
    \caption{A schematic diagram showing the reflected ($r$) and transmitted ($t$) waves when an incident ($i$) wave enters an interface of thickness $d$.}
    \label{fig:refr}
\end{figure}
%
Figure~\ref{fig:refr} shows the refraction and reflections from a simple two layer system, where layer \num{0} is a vacuum above the sample. 
It is clear, by analogy to Bragg's law, how the two waves, labelled $r$, could interfere constructively or destructively depending on the thickness of layer \num{1}, $d$.

@@

An heuristic explanation for the phase problem can be built in terms of the auto-correlation function of $\rmd\rho/\rmd z$, which gives a real-space representation of the information contained in the reflectivity pattern,
%
\begin{equation}
    \begin{aligned}
        \acf_{\rho'}(z) & = \int_{-\infty}^{\infty} \rho'(z)^* \, \rho'(z + \xi) \, \rmd \xi \\
                        & = \int_{-\infty}^{\infty} \left| \widehat{\rho'}(q)  \right|^{2} \rm{e}^{iqz} \rmd q,
    \end{aligned}
    \label{equ:acf}
\end{equation}
%
where the hat notation indicates a Fourier transform, the star notation a complex conjugate, and the prime notation a derivative. 
\begin{figure}
    \includegraphics[width=0.49\textwidth]{phase_problem_ACF.pdf}    
    \caption{The phase problem explanation: (a) the derivatives of the two hypothetical scattering length density profiles from Figure~\ref{fig:phase_problem_sld} are scaled $\delta$ functions peaking at the interfaces between scattering length density layers, (b) the autocorrelation functions, $\acf_{\rho'}(z)$ are almost identical between the samples, peaking at exactly the same locations.}
    \label{fig:phase_problem_acf}
\end{figure}

@@

The global optimisation of a reflectometry model is a particularly difficult problem, this is due to the ill-posed nature of this data, this is where there are many reasonable solutions to a particular reflectometry profile.
However, a particular global optimisation method has shown substantial utility in the fitting of reflectometry data \cite{varderlee_comparison_2007}, differential evolution \cite{wormington_characterization_1999}.
This has lead to the inclusion of this method in many common reflectometry analysis packages \cite{bjorck_fitting_2011}.

@@

The analysis of a reflectometry can take advantage of contrast variation, the exact methods of which are covered in other articles \cite{schurtenberger_contrast_2002}.
However, we note here that, the phrase ``global optimisation'' is often used to describe the analysis of multiple contrasts of the same structure. 
This is where a single structural model is used for the analysis of multiple measurements, but the scattering length densities are varied based on the particular contrasts under investigation. 
This method has been used to recover the phase information in neutron reflectometry analysis \cite{majkrzak_exact_1995,majkrzak_first_2000,majkrzak_phase_2003,koutsioubas_model_2019}.
However, it is also a tool in model-dependent analysis, as it reduces the number of local minimia in the parameter space. 
Typically, a global (here meaning multiple contrasts) optimisation criteria is used, for example the average or summed likelihood across all of the different measurements and model reflectometries.

@@

The Hessian matrix, $H$ describes the curvature of optimisation criteria, the likelihood, with respect to each of the model parameters, and it is frequently used in the gradient-based optimisation methods. 
This means it is an $i\times i$ matrix, where $i$ is again the number of model parameters, $\theta$, for a two parameter model, this would look like this,
%
\begin{equation}
    H = 
    \left[\begin{matrix}
        \dfrac{\partial^2 L}{\partial \theta_{1}^2} & \dfrac{\partial^2 L}{\partial \theta_1 \theta_2} \\[6pt]
        \dfrac{\partial^2 L}{\partial \theta_2 \theta_1} & \dfrac{\partial^2 L}{\partial \theta_{2}^2}
    \end{matrix}\right].
\end{equation}
%
If the parameter distribution is Gaussian in shape then the second derivative at the Gaussian maximum will give the variance of that distribution. 
Therefore, when these gradient approaches are used, the Hessian matrix is calculated and the diagonal of that matrix gives the uncertainty in each of the parameters. 
The off-diagonal elements contain information about the covariance between different parameters, however this is rarely leveraged and the covariances are taken to be negligible. 
Additionally, we note that often the Hessian is not calculated in an optimisation so the Jacobian matrix (the first derivative analogue of the Hessian) is used to estimate Hessian matrix.

The determination of parameter uncertainties from the Hessian matrix is dependent on two major assumptions about the parameter space:
\begin{enumerate}
    \item {the parameters are normally distributed with a single maximum,}
    \item {the covariance between the parameters is negligible.}
\end{enumerate}
However, this may not always be the case \cite{mccluskey_bayesian_2019}, therefore is it necessary to use methods to completely describe the parameter probability distribution.
One such method for this is Markov-chain Monte Carlo (MCMC), which samples the posterior probability distribution for each of our parameters to obtain an analytical description \cite{sivia_data_2006}.
Typically, MCMC is used on already optimised solutions to a particular problem, is in reflectometry analysis it is usually applied after the differential evolution has optimised the structure. 
In addition to being able to quantify the inverse uncertainties (the name given to the uncertainties in the model parameters) MCMC also offers a more complete understanding of the correlations between different parameters \cite{gilks_markov_1995}, which is particularly important in the ill-posed reflectometry analysis.

@@

Figure~\ref{fig:mcmc} shows the result of an MCMC sampling for a pair of overlapping Gaussian functions, performed using the \texttt{emcee} Python package \cite{foremanmackey_emcee_2012}.
The posterior probability distributions that are available to each of the four parameters are shown along with the data, the optimised fit and a subset of models within the posterior distributions. 
All of the samples from the posterior distributions fit within the uncertainty error bars on the data.
This shows the ability for MCMC to sample the range of the distribution that is allowed by the experimental uncertainty. 
%
\begin{figure}[t]
    \includegraphics[width=0.49\textwidth]{mcmc}
    \caption{An example of the result of MCMC uncertainty sampling on data synthesized from two overlapping Gaussian functions. $\theta_1$ and $\theta_2$ are the integrals of the Gaussian functions while $\theta_3$ and $\theta_4$ are the positions. The data is shown with blue circles, the optimised model with an orange line, and samples from the posterior distributions with green lines.}
    \label{fig:mcmc}
\end{figure}
%

@@

\section*{Acknowledgements}
ARM is supported by the Ada Lovelace Centre – a joint initiative between the Science and Technology Facilities Council (as part of UK Research and Innovation), Diamond Light Source, and the UK Atomic Energy Authority.
The authors would like to thank Luke Clifton of ISIS Neutron and Muon Source for insightful input on the manuscript.

>>>>

\begin{abstract}
Neutron reflectometry analysis is an inherently ill-posed problem, which is to say that there are many possible solutions which agree equally well with the measured data.
This leads to the use of model-dependent analysis to interpret the experimental data, where information that is ``known'' about the system is integrated into the analysis.
This tutorial aims to introduce the mathematics that underlies the use of model-dependent analysis in neutron reflectometry.
I hope that those that are well experienced in reflectometry find it as useful as those who are just about to experience their first reflectometry experiment.
\end{abstract}

@@

The Born approximation \cite{born_quantenmechanik_1926}, which assumes a single scattering event for each incident radiation particle, allows the reflected intensity to be calculated as follows \cite{sivia_elementary_2011},
%
\begin{equation}
    R(q) \approx \frac{16\pi^2}{q^4} \bigg| \int^{+\infty}_{-\infty}{\rho'(z)\exp{(-\mathrm{i} zq) \text{d}z} \bigg|^2},
    \label{equ:kine}
\end{equation}
%
where, $\rho'(z)$ is the first derivative of the scattering length density profile.
The scattering length density profile is our model, a functional description of how the scattering length density varies with depth normal to the interface, $z$.

@@

\section{Recursive methods}
The breakdown of the kinematic approach leads to the application of the Abel\`{e}s \cite{abeles_sur_1948} or Parratt \cite{parratt_surface_1954}, recursive model for the reflection of light at a given number of stratified interfaces, also referred to as the dynamical approach.
This is an alternative method for calculating the reflected intensity from a model, however, now the model is a series of layers of different scattering length density.
The probing radiation can be either reflected or refracted at the interface between each of the layers.
%
\begin{figure}[t]
    \includegraphics[width=\linewidth]{dynamic}
    \caption{A schematic diagram showing the reflected ($r$) and transmitted ($t$) waves when an incident ($i$) wave enters an interface of thickness $d$.}
    \label{fig:refr}
\end{figure}
%
Figure~\ref{fig:refr} shows the refraction and reflections from a simple two layer system, where layer \num{0} is a vacuum above the sample.
It is clear, by analogy to Bragg's law, how the two waves, labelled $r$, could interfere constructively or destructively depending on the thickness of layer \num{1}, $d$.

@@

A heuristic explanation for the phase problem can be built in terms of the auto-correlation function of $\rmd\rho/\rmd z$, which gives a real-space representation of the information contained in the reflectivity pattern,
%
\begin{equation}
    \begin{aligned}
        \acf_{\rho'}(z) & = \int_{-\infty}^{\infty} \rho'(z)^* \, \rho'(z + \xi) \, \rmd \xi \\
                        & = \int_{-\infty}^{\infty} \left| \widehat{\rho'}(q)  \right|^{2} \rm{e}^{iqz} \rmd q,
    \end{aligned}
    \label{equ:acf}
\end{equation}
%
where the hat notation indicates a Fourier transform, the star notation a complex conjugate, and the prime notation a derivative.
\begin{figure}
    \includegraphics[width=0.49\textwidth]{phase_problem_ACF.pdf}
    \caption{The phase problem explanation: (a) the derivatives of the two hypothetical scattering length density profiles from Figure~\ref{fig:phase_problem_sld} are scaled $\delta$ functions peaking at the interfaces between scattering length density layers, (b) the autocorrelation functions, $\acf_{\rho'}(z)$ are almost identical between the samples, peaking at exactly the same locations.}
    \label{fig:phase_problem_acf}
\end{figure}

@@

The global optimisation of a reflectometry model is a particularly difficult problem, this is due to the ill-posed nature of this data, this is where there are many reasonable solutions to a particular reflectometry profile.
However, a particular global optimisation method has shown substantial utility in the fitting of reflectometry data \cite{varderlee_comparison_2007}: differential evolution \cite{wormington_characterization_1999}.
This has lead to the inclusion of this method in many common reflectometry analysis packages \cite{bjorck_fitting_2011}.

@@

The analysis of a reflectometry model can take advantage of contrast variation, the exact methods of which are covered in other articles \cite{schurtenberger_contrast_2002}.
However, we note here that, the phrase ``global optimisation'' is often used to describe the analysis of multiple contrasts of the same structure.
This is where a single structural model is used for the analysis of multiple measurements, but the scattering length densities are varied based on the particular contrasts under investigation.
This method has been used to recover the phase information in neutron reflectometry analysis \cite{majkrzak_exact_1995,majkrzak_first_2000,majkrzak_phase_2003,koutsioubas_model_2019}.
However, it is also a tool in model-dependent analysis, as it reduces the number of local minimia in the parameter space.
Typically, a global (here meaning multiple contrasts) optimisation criteria is used. For example, the average or summed likelihood across all of the different measurements and model reflectometries.

@@

The Hessian matrix, $H$ describes the curvature of optimisation criteria, the likelihood, with respect to each of the model parameters, and it is frequently used in the gradient-based optimisation methods.
This means it is an $i\times i$ matrix, where $i$ is again the number of model parameters, $\theta$. For a two parameter model, this would look like,
%
\begin{equation}
    H =
    \left[\begin{matrix}
        \dfrac{\partial^2 L}{\partial \theta_{1}^2} & \dfrac{\partial^2 L}{\partial \theta_1 \theta_2} \\[6pt]
        \dfrac{\partial^2 L}{\partial \theta_2 \theta_1} & \dfrac{\partial^2 L}{\partial \theta_{2}^2}
    \end{matrix}\right].
\end{equation}
%
If the parameter distribution is Gaussian in shape then the second derivative at the Gaussian maximum will give the variance of that distribution.
Therefore, when these gradient approaches are used, the Hessian matrix is calculated and the diagonal of that matrix gives the uncertainty in each of the parameters.
The off-diagonal elements contain information about the covariance between different parameters, however this is rarely leveraged and the covariances are taken to be negligible.
Additionally, we note that often the Hessian is not calculated in an optimisation so the Jacobian matrix (the first derivative analogue of the Hessian) is used to estimate Hessian matrix.

The determination of parameter uncertainties from the Hessian matrix is dependent on two major assumptions about the parameter space:
\begin{enumerate}
    \item {the parameters are normally distributed with a single maximum,}
    \item {the covariance between the parameters is negligible.}
\end{enumerate}
However, this may not always be the case \cite{mccluskey_bayesian_2019}. Therefore, is it necessary to use methods to completely describe the parameter probability distribution.
One such method for this is Markov-chain Monte Carlo (MCMC), which samples the posterior probability distribution for each of our parameters to obtain an analytical description \cite{sivia_data_2006}.
Typically, MCMC is used on already optimised solutions to a particular problem. In reflectometry analysis it is usually applied after the differential evolution has optimised the structure.
In addition to being able to quantify the inverse uncertainties (the name given to the uncertainties in the model parameters) MCMC also offers a more complete understanding of the correlations between different parameters \cite{gilks_markov_1995}, which is particularly important in the ill-posed reflectometry analysis.

@@

Figure~\ref{fig:mcmc} shows the result of a MCMC sampling for a pair of overlapping Gaussian functions, performed using the \texttt{emcee} Python package \cite{foremanmackey_emcee_2012}.
The posterior probability distributions that are available to each of the four parameters are shown along with the data, the optimised fit and a subset of models within the posterior distributions.
All of the samples from the posterior distributions fit within the uncertainty error bars on the data.
This shows the ability for MCMC to sample the range of the distribution that is allowed by the experimental uncertainty.
%
\begin{figure}[t]
    \includegraphics[width=0.49\textwidth]{mcmc}
    \caption{An example of the result of MCMC uncertainty sampling on data synthesized from two overlapping Gaussian functions. $\theta_1$ and $\theta_2$ are the integrals of the Gaussian functions while $\theta_3$ and $\theta_4$ are the positions. The data is shown with blue circles, the optimised model with an orange line, and samples from the posterior distributions with green lines.}
    \label{fig:mcmc}
\end{figure}
%

@@

\section*{Acknowledgements}
ARM is supported by the Ada Lovelace Centre -- a joint initiative between the Science and Technology Facilities Council (as part of UK Research and Innovation), Diamond Light Source, and the UK Atomic Energy Authority.
The authors would like to thank Luke Clifton of ISIS Neutron and Muon Source for insightful input on the manuscript.
